Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2020-12-28 article.tex
Misc changes

	A mutation with a fitness benefit should have a tendency to fix, while detrimental mutations fix less often.
	Fixation and loss of a mutation is, however, not only determined by its intrinsic fitness benefit, but also by the fitness of the background it arose on and the fitness of competing variants \citep{strelkowa_clonal_2012}.
	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
	These interference effects, along with stochastic epidemiological dynamics and transient selection, limit the predictability of evolution.
	Here, we investigate the frequency trajectories of amino acid mutations and the degree to which fixation or loss are predictable.
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
	To our surprise, we find that the predictability of these trajectories is very limited: for A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions.
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retain clear signatures of predictability.
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

For $f_0=0.3$, we found $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{B} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. The same analysis was performed for A/H1N1pdm, with the $89$ found trajectories displayed in figure \sref{fig:trajectories_h1n1}.
Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time.
This is due to the fact that some trajectories ``skipped" the interval $f_0$ in question on their initial rise due to sparse sampling.
These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$.
Removing them does not change results significantly.

@@

To quantify the extent to which this preconception of sweeping adaptive mutations is true, we estimated the probability distribution $P_{\Delta t}(f\vert\,f_0)$ of finding a trajectory at frequency $f$ after a time $\Delta t$ given that it was observed at $f_0$ at time $0$.
The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}C.
Initially, \emph{i.e.}, at time $t_0=0$, this distribution is by construction peaked around $f_0$.
If a large fraction of the trajectories keep increasing after  this time, we should see the ``mass'' of $P_{\Delta t}(f\vert\,f_0)$ move to the right towards higher frequencies as time progresses.

However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps.
The thick black line in Figure~\ref{fig:trajectories}B shows the average frequency of all trajectories.
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (see supplement).
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$. \textbf{C}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{B} and for specific values of $\Delta t$. }
	\label{fig:trajectories}
\end{figure*}

@@

@@

This near equivalence of the consensus and the strain with highest LBI can be explained as follows:
The LBI tends to be high for nodes in a tree that are close to the root of a dense and large clade.
A typical sample of influenza HA sequences fall into a small number of recognizable clades, and the strains with maximal LBI will often be close to the root of the largest of those clades.
This root of the largest clade will often be close to the consensus of the whole population, explaining the similar distance patterns.
To test that hypothesis, we measure the hamming distance from the sequence of the top LBI strain to the consensus sequence for populations of all time bins.
Panel \textbf{B} of figure~\ref{fig:LBI_consensus} shows these distances, scaled with respect to an average strain (details in caption). It clearly shows that the top-LBI strain and the consensus sequence are indeed quite similar: out of 48 time bins, only once is the sequence of the top-LBI strain farther away from the consensus than the average sequence is. Moreover, the sequence of the top-LBI strain \emph{exactly} matches the consensus in 19 cases.

@@

A recent preprint proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
Furthermore, selection might cause the rapid rise of a novel variant to macroscopic frequencies (observable in a global sample) but its benefit rapidly ``expires'' because competing variants catch up and/or it mediates immune escape only to a small fraction of the population.
These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.

@@

\subsection*{Data and code availability} % (fold)
\label{sub:data}
	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}.
	Strain names and accession numbers are given as tables in two supplementary files. Outliers strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/src/config} were removed. \\
	The code used to generate the figures presented here is available at \url{https://github.com/PierreBarrat/FluPredictibility}.
	% https://github.com/nextstrain/seasonal-flu/tree/081be2f487182252627418491be36cd01d645954/config
% subsection data (end)

>>>>

	A mutation with a fitness benefit should have a tendency to fix, while detrimental mutations fix less often.
	Fixation and loss of a mutation is, however, not only determined by its intrinsic fitness benefit, but also by the fitness of the background it arose on and the fitness of competing variants \citep{strelkowa_clonal_2012}.
	This degree to which these different components affect influenza evolution has been investigated by \citet{illingworth_components_2012}, who showed that intrinsic selection, background fitness, and later interference have comparable effects on fixation.
	These interference effects, along with stochastic epidemiological dynamics and transient selection, limit the predictability of evolution.
	Here, we investigate the frequency trajectories of amino acid mutations and the degree to which fixation or loss are predictable.
	We quantify how rapidly mutations at different frequencies are lost or fixed and how rapidly they spread through the population.
	We further investigate whether any properties or statistics are predictive of whether a particular mutation fixes or not.
	To our surprise, we find that the predictability of these trajectories is very limited: for A/H3N2, the probability that a mutation fixes differs little from its current frequency, as would be expected if fixation happened purely by chance.
	This observation holds for many different categories of mutations, including mutations at epitope positions.
	This weak predictability is not attributable solely to clonal interference and genetic linkage, as simulation of models including even strong interference retains clear signatures of predictability.
	Consistent with these observations, we show that a simple predictor uninformed by fitness, the consensus sequence, performs as the well as the Local Branching Index (LBI), the growth measure based on the genealogy used in \cite{neher_predicting_2014}.
	This suggests that although LBI has predictive power, the reason for its success may not be related to it approximating fitness of strains.

@@

For $f_0=0.3$, we found $120$ such trajectories in the case of A/H3N2 influenza, represented on the panel \textbf{C} of figure~\ref{fig:trajectories}, where time is shifted such that $t_0 = 0$. The same analysis was performed for A/H1N1pdm, with the $89$ found trajectories displayed in figure \sref{fig:trajectories_h1n1}.
Some trajectories fall in the frequency bin around $f_0$ while decreasing, even though they crossed that bin at an earlier time.
This is due to the fact that some trajectories ``skipped" the interval $f_0$ in question on their initial rise due to sparse sampling.
These trajectories are nevertheless rising in the sense that they start at frequency 0 for $t\rightarrow -\infty$.
Removing them does not change results significantly.

@@

To quantify the extent to which this preconception of sweeping adaptive mutations is true, we estimated the probability distribution $P_{\Delta t}(f\vert\,f_0)$ of finding a trajectory at frequency $f$ after a time $\Delta t$ given that it was observed at $f_0$ at time $0$.
The results for different $\Delta t$ are shown in figure~\ref{fig:trajectories}B.
Initially, \emph{i.e.}, at time $t_0=0$, this distribution is by construction peaked around $f_0$.
If a large fraction of the trajectories keep increasing after  this time, we should see the ``mass'' of $P_{\Delta t}(f\vert\,f_0)$ move to the right towards higher frequencies as time progresses.

However, future distributions for $\Delta t >0$ do not seem to follow a pattern compatible with selective sweeps.
The thick black line in Figure~\ref{fig:trajectories}C shows the average frequency of all trajectories.
This average makes a sharp turn at $t=0$ and is essentially flat for $t>0$ in the case of A/H3N2, and slightly increasing for A/H1N1pdm (figure \sref{fig:trajectories_h1n1}).
Hence, the fact that this average rose for $t<0$ gives little information for $t>0$, and is due to the conditions by which these trajectories were selected.
This shows that sweep-like trajectories rising steadily from frequency $0$ to $1$ are not common enough to dominate the average trajectory.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{C} and for specific values of $\Delta t$. \textbf{C}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$.}
	\label{fig:trajectories}
\end{figure*}

@@

Next, we cluster trajectories of simulated populations based on the intrinsic fitness effect of corresponding mutations and in the case of a fast changing fitness landscape ($\Delta t=10$). 
Since competition between adaptive mutations and genetic linkage is very strong in this regime, we do not expect the intrinsic fitness effect to be fully predictive of fixation. 
However, figure \sref{fig:simulations_fitness} shows that mutations with an above median fitness effect have a significantly larger chance to fix than those with a below median fitness effect. 
This strongly contrasts with figure \ref{fig:pfix_stratified}, and confirms the intuition that intrinsic fitness effect can be partly predictive of fixation even in a regime of strong clonal interference.

@@

This near equivalence of the consensus and the strain with highest LBI can be explained as follows:
The LBI tends to be high for nodes in a tree that are close to the root of a dense and large clade.
A typical sample of influenza HA sequences fall into a small number of recognizable clades, and the strains with maximal LBI will often be close to the root of the largest of those clades.
This root of the largest clade will often be close to the consensus of the whole population, explaining the similar distance patterns.
To test that hypothesis, we measure the hamming distance from the sequence of the top LBI strain to the consensus sequence for populations of all time bins.
Panel \textbf{B} of figure~\ref{fig:LBI_consensus} shows these distances, scaled with respect to an average strain (details in caption). It clearly shows that the top-LBI strain and the consensus sequence are indeed quite similar: out of 48 time bins, only twice is the sequence of the top-LBI strain farther away from the consensus than the average sequence is. Moreover, the sequence of the top-LBI strain \emph{exactly} matches the consensus in 19 cases.

@@

A recent study proposed that influenza virus evolution is primarily limited by an asynchrony between population level selection and generation of new variants within infected hosts \citep{morris_asynchrony_2020}.
Along these lines, it is possible that the A/H3N2 population readily responds once population level selection is high enough by giving rise to essentially equivalent variants.
Furthermore, selection might cause the rapid rise of a novel variant to macroscopic frequencies (observable in a global sample) but its benefit rapidly ``expires'' because competing variants catch up and/or it mediates immune escape only to a small fraction of the population.
These considerations might explain the disconnect between models of rapid adaptation and the frequency dynamics observed in influenza virus populations.

@@

\subsection*{Data and code availability} % (fold)
\label{sub:data}
	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}.
	Strain names and accession numbers are given as tables in two supplementary files. Outliers strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/tree/master/src/config} were removed. \\
	The code used to generate the figures presented here is available at \url{https://github.com/PierreBarrat/FluPredictibility}.
	% https://github.com/nextstrain/seasonal-flu/tree/081be2f487182252627418491be36cd01d645954/config
% subsection data (end)
